### 2018-03-13 submission/manuscript.Rmd
move problem section to full taxa vs combined taxa section

***Combined taxa model classifies subjects better than using individual taxa.*** 
Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models. For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`). For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2]. Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study. Within the significant OR taxa only models the most important taxa across study for stool were *Ruminococcus* and *Clostridium XI* based on mean decrease in accuracy [Figure 4A]. Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for *Blautia* and *Weissella* (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).  For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3]. For the significant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the most important based on mean decrease in accuracy [Figure 4B]. Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.

***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs.*** 
Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR. Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (`r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)(min-max)) [Figure 5A & S4A]. In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S4B-C]. When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed. Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2))`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%). The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Although the significant OR taxa models can classify individuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.

>>>>

***Combined taxa model classifies subjects better than using individual taxa.*** 
Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models. For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).  Within the significant OR taxa only models the most important taxa across study for stool were *Ruminococcus* and *Clostridium XI* based on mean decrease in accuracy [Figure 4A]. Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for *Blautia* and *Weissella* (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`). For the significant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the most important based on mean decrease in accuracy [Figure 4B]. Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.

***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs.*** 
Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR. Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (`r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)(min-max)) [Figure 5A & S2A]. In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S2B-C]. When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed. Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2))`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%). The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). For these full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S3]. Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S4]. Although the significant OR taxa models can classify individuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.

### 2018-03-13 submission/manuscript.tex
move problem section to full taxa vs combined taxa section

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa.}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from stool samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). For the full
taxa models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae}
were the most common taxa in the top 10\% mean decrease in accuracy
(MDA) across studies {[}Figure S2{]}. Of the 3 taxa with significant
ORs, all 3 were among the top 10\% most important taxa as measured by
mean decrease in accuracy, in at least one study. Within the significant
OR taxa only models the most important taxa across study for stool were
\emph{Ruminococcus} and \emph{Clostridium XI} based on mean decrease in
accuracy {[}Figure 4A{]}. Similarly, using data from the unmatched
tissue samples, the combined model had an AUC of 0.77, which was
significantly higher than the AUC values for \emph{Blautia} and
\emph{Weissella} (P-value \textless{} 0.037). For the full taxa models
using unmatched tissue, \emph{Lachnospiraceae}, \emph{Bacteroidaceae},
and \emph{Ruminococcaceae} were the most common taxa in the top 10\%
mean decrease in accuracy across studies {[}Figure S3{]}. For the
significant OR taxa unmatched tissue models both \emph{Dorea} and
\emph{Blautia} were the most important based on mean decrease in
accuracy {[}Figure 4B{]}. Pooling the information from the taxa with
significant ORs results in a model that outperforms classifications made
using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs.}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were only marginally better than 0.5 for any study (median AUC = 0.549
(0.367 - 0.971)(min-max)) {[}Figure 5A \& S4A{]}. In contrast, the
models for classifying subjects as having normal colons or having
carcinomas using data from stool or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 5B \& S4B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from stool samples we found that the AUC for 6 of 7 studies increased by
an average of 14.8\%) and AUC for the Flemer study decreased by 0.54\%).
The overall improvement in performance was statistically significant
(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). Similarly,
using the data from unmatched tissue samples we found that the AUC of
studies increased by an average of 19.11\% when we used all of the taxa
rather than the 3 taxa with significant ORs (one-tailed paired T-test;
P-value = 0.03). Although the significant OR taxa models can classify
individuals with and without carcinoma tumors, they are still missing
taxa from the full community models that can increase the model
accuracy.

>>>>

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa.}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from stool samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). Within the
significant OR taxa only models the most important taxa across study for
stool were \emph{Ruminococcus} and \emph{Clostridium XI} based on mean
decrease in accuracy {[}Figure 4A{]}. Similarly, using data from the
unmatched tissue samples, the combined model had an AUC of 0.77, which
was significantly higher than the AUC values for \emph{Blautia} and
\emph{Weissella} (P-value \textless{} 0.037). For the significant OR
taxa unmatched tissue models both \emph{Dorea} and \emph{Blautia} were
the most important based on mean decrease in accuracy {[}Figure 4B{]}.
Pooling the information from the taxa with significant ORs results in a
model that outperforms classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs.}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were only marginally better than 0.5 for any study (median AUC = 0.549
(0.367 - 0.971)(min-max)) {[}Figure 5A \& S2A{]}. In contrast, the
models for classifying subjects as having normal colons or having
carcinomas using data from stool or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 5B \& S2B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from stool samples we found that the AUC for 6 of 7 studies increased by
an average of 14.8\%) and AUC for the Flemer study decreased by 0.54\%).
The overall improvement in performance was statistically significant
(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). For these
full taxa models using stool, \emph{Bacteroides} and
\emph{Lachnospiraceae} were the most common taxa in the top 10\% mean
decrease in accuracy (MDA) across studies {[}Figure S3{]}. Similarly,
using the data from unmatched tissue samples we found that the AUC of
studies increased by an average of 19.11\% when we used all of the taxa
rather than the 3 taxa with significant ORs (one-tailed paired T-test;
P-value = 0.03). For the full taxa models using unmatched tissue,
\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
decrease in accuracy across studies {[}Figure S4{]}. Although the
significant OR taxa models can classify individuals with and without
carcinoma tumors, they are still missing taxa from the full community
models that can increase the model accuracy.
